KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-04-05 paper_overleaf/main.tex
Updates from Overleaf

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya (Laurits et al., 2022, in press). The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (Laurits et al., 2022, in press). All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

@@

Contamination can affect the accuracy of relatedness inference for two reasons. For one, if a substantial fraction of samples is contaminated, then the estimation of $p_0$ becomes inaccurate, because the majority of pairwise differences will include at least one contaminated sample. The second issue is that contaminated samples will look less similar to other individuals, and thus cause a bias towards inferring them to be less related to other individuals. The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.  Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way (Laurits et al., 2022 (in press)). SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available. However, the decrease in power is lower for KIN compared to READ.

The Lech valley data  has low contamination, and no ROH. For pairwise comparisons with large numbers of overlapping sites (>10000), KIN, READ and lcMLKIN all mostly agree. However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples. We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (Laurits et al., 2022 (in press)). In addition, KIN identifies a pair of third degree relatives. In this case of a population with large amounts of ROH, we find that the inference by lcMLKIN are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.

>>>>

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}. The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}. All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans \cite{laurits_skov_genetic_nodate}, and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

@@

Contamination can affect the accuracy of relatedness inference for two reasons. For one, if a substantial fraction of samples is contaminated, then the estimation of $p_0$ becomes inaccurate, because the majority of pairwise differences will include at least one contaminated sample. The second issue is that contaminated samples will look less similar to other individuals, and thus cause a bias towards inferring them to be less related to other individuals. The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.  Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}. SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available. However, the decrease in power is lower for KIN compared to READ.

The Lech valley data  has low contamination, and no ROH. For pairwise comparisons with large numbers of overlapping sites (>10000), KIN, READ and lcMLKIN all mostly agree. However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples. We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}. In addition, KIN identifies a pair of third degree relatives. In this case of a population with large amounts of ROH, we find that the inference by lcMLKIN are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
